---
title: "`r params$project_name`"
subtitle: "Single-cell RNA-seq analysis report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
params:
  project_name: "Single-cell RNA-seq analysis"
  outdir: "results"
  scripts_dir: "."
  figures_dir: "results/figures"
  tables_dir: "results/tables"
  metrics_dir: "results/metrics"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    self_contained: true
    theme: flatly
  pdf_document:
    toc: true
    toc_depth: 2
    latex_engine: tectonic
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      out.width = "100%", fig.align = "center")
suppressPackageStartupMessages({ library(knitr) })

figdir <- params$figures_dir
tabdir <- params$tables_dir

read_tsv_safe <- function(name) {
    p <- file.path(tabdir, paste0(name, ".tsv"))
    if (!file.exists(p)) return(NULL)
    out <- tryCatch(read.delim(p, check.names = FALSE), error = function(e) NULL)
    if (is.null(out) || nrow(out) == 0) return(NULL)
    out
}
fig_exists <- function(name) file.exists(file.path(figdir, paste0(name, ".png")))

# Figures are produced by the Python stages and live outside the render directory.
# By default knitr rewrites an image path to be relative to the output document, which
# breaks when the figures and the report sit on different branches of the tree.
# rel_path = FALSE keeps the absolute path, which both the self-contained HTML
# (base64-embedded at render time) and the LaTeX/PDF path resolve correctly.
show_fig <- function(name, caption = NULL) {
    src <- file.path(figdir, paste0(name, ".png"))
    if (!file.exists(src)) {
        cat("\n*Figure", name, "was not produced for this run.*\n\n")
        return(invisible(NULL))
    }
    knitr::include_graphics(normalizePath(src), rel_path = FALSE)
}
note_missing <- function(what) {
    cat("\n*", what, "was not available for this run - this section is omitted.*\n\n")
}

# Each stage writes its own <stage>.json; merge them here. A stage that did not run
# simply has no file, and get_metric() then returns the supplied default.
metrics <- list()
if (dir.exists(params$metrics_dir) && requireNamespace("jsonlite", quietly = TRUE)) {
    for (f in list.files(params$metrics_dir, pattern = "\\.json$", full.names = TRUE)) {
        m <- tryCatch(jsonlite::fromJSON(f, simplifyVector = TRUE),
                      error = function(e) NULL)
        if (!is.null(m)) metrics <- utils::modifyList(metrics, m)
    }
}
get_metric <- function(section, key, default = NA) {
    v <- tryCatch(metrics[[section]][[key]], error = function(e) NULL)
    if (is.null(v) || length(v) == 0) return(default)
    if (length(v) > 1) return(paste(v, collapse = ", "))
    v
}
```

# Overview

```{r overview}
ov <- data.frame(
    Item = c("Samples", "Cells before QC", "Cells after QC", "Cells analysed",
             "Genes analysed", "Doublets removed", "Clusters",
             "Batch integration", "Marker genes"),
    Value = c(
        get_metric("qc", "n_samples"),
        get_metric("qc", "cells_before"),
        get_metric("qc", "cells_after"),
        get_metric("cluster", "cells"),
        get_metric("cluster", "n_hvg"),
        get_metric("doublets", "removed"),
        get_metric("cluster", "n_clusters"),
        get_metric("cluster", "integration"),
        get_metric("markers", "n_markers")
    ), stringsAsFactors = FALSE)
ov$Value <- as.character(ov$Value)
ov <- ov[!is.na(ov$Value) & ov$Value != "NA", ]
kable(ov, row.names = FALSE, caption = "Analysis summary")
```

Cells were quality-filtered, screened for doublets, normalised and clustered following
the Scanpy workflow described in the
[single-cell best-practices book](https://www.sc-best-practices.org/). Each section below
states the thresholds actually used, and the tables behind every figure are in
`tables/`.

# Quality control

```{r qc-thresh}
th <- tryCatch(metrics[["qc"]][["thresholds"]], error = function(e) NULL)
if (!is.null(th) && length(th)) {
    d <- data.frame(Threshold = names(th),
                    Value = as.character(unlist(lapply(th, function(x)
                        if (is.null(x) || length(x) == 0) "none" else as.character(x)))),
                    stringsAsFactors = FALSE)
    kable(d, row.names = FALSE, caption = "Cell/gene filtering thresholds applied")
}
```

```{r qc-before}
show_fig("qc_before")
```

Distributions **before** filtering: genes detected per cell, total UMI counts per cell,
and the percentage of counts from mitochondrial genes. Low gene counts indicate empty or
dying droplets; a high mitochondrial fraction indicates stressed or lysing cells; very
high counts can indicate doublets.

```{r qc-after}
show_fig("qc_after")
```

```{r qc-tables}
before <- read_tsv_safe("qc_per_sample_before")
after <- read_tsv_safe("qc_per_sample_after")
if (!is.null(before)) kable(before, row.names = FALSE, digits = 1,
                            caption = "Per sample, before filtering")
if (!is.null(after)) kable(after, row.names = FALSE, digits = 1,
                           caption = "Per sample, after filtering")
filt <- read_tsv_safe("qc_filtering")
if (!is.null(filt)) kable(filt, row.names = FALSE,
                          caption = "Cells/genes removed, by reason")
```

`r if (identical(as.character(get_metric("qc","mito_genes_found")), "0")) "**Note:** no mitochondrial genes were found in this dataset, so the mitochondrial-fraction filter removed nothing. This is expected when gene identifiers are Ensembl IDs rather than symbols."`

# Doublet detection

```{r doublets}
if (identical(as.character(get_metric("doublets", "enabled")), "FALSE")) {
    cat("\n*Doublet detection was disabled for this run.*\n\n")
} else {
    show_fig("doublet_scores")
    d <- read_tsv_safe("doublets_per_sample")
    if (!is.null(d)) kable(d, row.names = FALSE, digits = 2,
                           caption = "Scrublet results per sample")
}
```

Doublets — droplets that captured two cells — otherwise appear as artificial
"intermediate" clusters that can be mistaken for transitional cell states. Scrublet
simulates doublets from the observed data and scores each barcode against them. Scoring
is done per sample, because doublet rate depends on how heavily each library was loaded.

# Clustering and structure

```{r umap-clusters}
show_fig("umap_leiden")
```

```{r cluster-params}
cp <- data.frame(
    Parameter = c("Highly variable genes", "Principal components", "Neighbours",
                  "Leiden resolution", "Clusters found", "Integration"),
    Value = as.character(c(
        get_metric("cluster", "n_hvg"), get_metric("cluster", "n_pcs"),
        get_metric("cluster", "n_neighbors"), get_metric("cluster", "resolution"),
        get_metric("cluster", "n_clusters"), get_metric("cluster", "integration"))),
    stringsAsFactors = FALSE)
kable(cp[!is.na(cp$Value) & cp$Value != "NA", ], row.names = FALSE,
      caption = "Clustering parameters")
```

Cluster count is a direct consequence of the Leiden `resolution` parameter — it is a
granularity dial, not a discovered quantity. Raising it splits clusters, lowering it
merges them. Treat clusters as a working partition to be validated against marker genes
and known biology, not as an objective number of cell types.

```{r umap-sample}
show_fig("umap_sample")
```

```{r integration-note, results='asis'}
integ <- as.character(get_metric("cluster", "integration"))
if (identical(integ, "none")) {
    cat("\nBatch integration was **not** applied. If the UMAP above separates by sample ",
        "rather than by cell type, that separation may be technical; consider setting ",
        "`integration.method: harmony`.\n\n", sep = "")
} else {
    cat("\nBatches were integrated with **", integ, "**. Integration can also remove ",
        "genuine biological differences between samples, so compare against an ",
        "unintegrated run before drawing conclusions about sample-specific populations.\n\n",
        sep = "")
}
```

```{r qc-umaps}
for (f in c("umap_pct_counts_mt", "umap_n_genes_by_counts", "umap_doublet_score")) {
    if (fig_exists(f)) show_fig(f)
}
```

```{r composition}
show_fig("cluster_composition")
comp <- read_tsv_safe("cluster_sizes")
if (!is.null(comp)) kable(comp, row.names = FALSE, caption = "Cells per cluster")
```

```{r pca-var}
show_fig("pca_variance")
```

# Marker genes

```{r markers}
m <- read_tsv_safe("markers")
if (is.null(m)) {
    note_missing("Marker gene results")
} else {
    top <- do.call(rbind, lapply(split(m, m$cluster), function(d) head(d, 5)))
    cols <- intersect(c("cluster", "gene", "logfoldchange", "pval_adj"), colnames(top))
    kable(top[, cols], row.names = FALSE, digits = c(0, 0, 2, 6),
          caption = "Top 5 marker genes per cluster")
}
```

```{r marker-figs}
show_fig("marker_dotplot")
show_fig("marker_heatmap")
```

Markers were identified with a Wilcoxon rank-sum test comparing each cluster against all
others. **These p-values are computed on the same cells used to define the clusters**, so
they are anti-conservative by construction and should be read as a ranking rather than as
evidence that a cluster is statistically real. The full table is in
`tables/markers_all.tsv`.

```{r marker-umaps}
gene_figs <- list.files(figdir, pattern = "^umap_gene_.*\\.png$")
for (f in head(gene_figs, 6)) show_fig(sub("\\.png$", "", f))
```

# Cell type annotation

```{r annotation}
ann <- read_tsv_safe("cluster_annotation")
scores <- read_tsv_safe("cell_type_scores")
if (is.null(ann)) {
    cat("\n*No marker sets were configured, so clusters are left unlabelled.* ",
        "Add `annotation.marker_sets` to the config to get provisional labels, or ",
        "annotate manually from the marker genes above.\n\n", sep = "")
} else {
    kable(ann, row.names = FALSE, caption = "Provisional cluster labels")
    show_fig("umap_cell_type")
    show_fig("cell_type_scores")
    if (!is.null(scores)) kable(scores, row.names = FALSE, digits = 3,
                                caption = "Mean marker-set score per cluster")
    cat("\nLabels are assigned by whichever supplied marker set scores highest in each ",
        "cluster, and are marked *low confidence* when the top two sets score nearly ",
        "equally. They are a starting hypothesis, only ever as good as the marker sets ",
        "provided, and should be confirmed against the marker genes and known biology ",
        "before being reported.\n\n", sep = "")
}
```

# Methods and provenance

```{r provenance}
prov <- data.frame(
    Item = c("Report generated", "Workflow", "Filtering", "Doublets",
             "Normalisation", "Clustering", "Markers", "Output directory"),
    Value = c(
        format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
        "Scanpy, following sc-best-practices.org",
        sprintf("min_genes=%s, max_pct_mt=%s",
                get_metric("qc", "cells_after", ""), get_metric("qc", "mito_genes_found", "")),
        if (identical(as.character(get_metric("doublets", "enabled")), "FALSE"))
            "not run" else "Scrublet, per sample",
        "shifted log (normalize_total + log1p)",
        sprintf("Leiden, resolution %s", get_metric("cluster", "resolution")),
        sprintf("Wilcoxon rank-sum (%s significant)", get_metric("markers", "n_markers")),
        normalizePath(params$outdir, mustWork = FALSE)),
    stringsAsFactors = FALSE)
kable(prov, row.names = FALSE, caption = "Provenance")
```

The annotated `AnnData` object is at `h5ad/04_annotated.h5ad` and can be loaded directly
for further analysis:

```
import anndata as ad
adata = ad.read_h5ad("h5ad/04_annotated.h5ad")
```

**Interpretation caveats.** Cluster count depends on the resolution parameter and is not
an objective property of the data. Marker gene p-values are inflated by double-dipping
(clustering and testing on the same cells). Cell-type labels here are computational
predictions, not ground truth. Differences in cluster proportions between samples are
*not* tested for significance in this report — with few biological replicates, apparent
compositional shifts are frequently noise, and testing them properly requires a method
that accounts for the compositional nature of the data.
